---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data
To get started, I loaded some libraries to assist in data analysis.

1. `dplyr` provides handy data manipulation tools and the `%>%` pipe operator.
2. `ggplot2` makes it easy to create modern charts.

```{r echo = T}
  require(dplyr,warn.conflicts=F)
  require(ggplot2,warn.conflicts=F)
```


Then I downloaded the data, checking to see if it the data already exists on disk. When the data is available, I unzipped it, again checking to make sure it hasn't already been done.

```{r echo = T}

  if( !file.exists('activity.zip')) {
    download.file(  'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
                  , destfile = 'activity.zip'
                  , method = 'curl')
    }
  if( !file.exists('activity.csv')) {
    unzip('activity.zip')
    }
```

Next, I read all the data into a single object, activity, then converted the date field to from a character class to POSIXct.

```{r echo = T}
  if( !exists('activity')) {
    activity  <- read.csv('activity.csv')
    activity$date  <- as.POSIXct(activity$date)
    }
```


## What is mean total number of steps taken per day?

To determine the total number of steps taken per day, I grouped the data by date and calculated some summary statistics: 

* total number of steps
* mean number of steps
* median number of steps

```{r echo = T}
  stepmeans  <- activity %>%
    group_by(date) %>%
    summarize(  totalsteps  = sum   (steps, na.rm = T)
              , meansteps   = mean  (steps, na.rm = T)
              , mediansteps = median(steps, na.rm = T)
      )
```

With that data in hand, I looked at the total number of steps per day with a quick histogram, developing a picture of the frequencies:
```{r fig.height=5, fig.width=7, echo= T}
  hist(stepmeans$totalsteps,xlab = 'Total Steps',col='blue',)
```

Behind that histogram is the calcultaed data itself. Below, I'm also including median for the sake of clarity:
```{r echo = T}
  stepmeans
```


## What is the average daily activity pattern?

To determine if any sorts of patterns exists by time, I grouped the data by time interval, then summarized it to calculate the mean number of steps taken each interval.

```{r echo = T}
  pattern  <- activity %>%
    group_by(interval) %>%
    summarize(meansteps = mean(steps,na.rm=T)) %>%
    arrange(interval)
```

I ploted the mean number of steps on the y axis against the time interval on the x axis as a line plot to get a high level overview of daily patterns.
```{r fig.height = 5, fig.width=7, echo = T}
  plot(meansteps ~ interval, data=pattern,type='l',ylab='mean steps',xlab='interval')
```

It seems that the interval with the most amount of steps is 8:35 am:
```{r echo = T}
  pattern %>%
    arrange(-meansteps) %>%
    head(1)
```

  

## Imputing missing values
Unfortunately, much of the data in this set is missing:
```{r echo = T}
  #number of rows with missing data
  sum(!complete.cases(activity))

  #percentage of data missing
  sum(!complete.cases(activity))/nrow(activity)
```

2304 rows with missing data means that over 13% of our data is missing.

However, only the step data is missing--date and interval data is still available. There is not a distinguishable pattern for the number of steps take per day, whereas there is a clear pattern for the time interval. With that in mind, I imputed missing data with the mean value of steps for the each interval.

```{r echo = T}
  a2  <- activity %>%
    left_join(pattern, by='interval') %>%
    mutate( imputedsteps = ifelse(is.na(steps),meansteps,steps)) %>%
    select( steps = imputedsteps,date,interval)
```

After I imputed any missing data, I recalulated the total, mean and median steps by day.
```{r echo = T}
  stepmeans2  <- activity %>%
    group_by(date) %>%
    summarize(  imputedtotalsteps  = sum   (steps, na.rm = T)
              , imputedmeansteps   = mean  (steps, na.rm = T)
              , imputedmediansteps = median(steps, na.rm = T)
      )
```

And then compared the new totals to the old with a histogram. Below, the blue data is the original histogram and the red figure displays the imputed data.
```{r fig.height=5, fig.width=7, echo = T}
  par(mfrow=c(1,2))
  hist(stepmeans$totalsteps,xlab = 'Total Steps',col='blue',main='Original')
  hist(stepmeans2$imputedtotalsteps,xlab = 'Imputed Steps',col='red',main='Imputed')
```

I also looked at the new means and median for each day, then compared them both to the original data to see what sort of impact the imputation had:

```{r echo=T}
changes  <- stepmeans2 %>%
  left_join(stepmeans,by='date') %>%
  select(   date
          , imputedmeansteps
          , imputedmediansteps
          , origmeansteps = meansteps
          , origmediansteps = mediansteps) %>%
  mutate(   changemean   = imputedmeansteps   - origmeansteps
          , changemedian = imputedmediansteps - origmediansteps)

summary(changes[,6:7])
```

Apparently, imputing new values based on the mean number of steps per interval induced no perceptible change to the data.


## Are there differences in activity patterns between weekdays and weekends?

To answer this question, I needed to add a new factor variable, determining whether the day was a weekday or a weekend.

```{r echo = T}
  a2  <- a2 %>%
    mutate( dayname = weekdays(date)
            , weekend = as.factor(ifelse(grepl("Sunday|Saturday",dayname),"weekend","weekday"))
            )
```

Next, I plotted the data showing the mean number of steps for each time interval with panels for weekends vs. weekdays.

```{r echo=T,fig.width=5,fig.height=7}
  a2 %>%
    group_by(interval,weekend) %>%
    summarize(meansteps = mean(steps)) %>%
    ggplot( aes(y=meansteps,x=interval)) + geom_line() + facet_wrap(~weekend)
```

The plots reveal a clear difference: 

* On week days, there's a large spike in the morning as people walk to begin their day and walking tapers off quickly before 8 pm. 
* On weekends, ramp up more slowly at the beginning of the day, are more consistent throughout the day, and carry on past 8 pm. 